# Coursera Reproducible Research: Course Project 2

## Severe weather events in the US and their impacts on health and economy

# Synopsis

In this report, we aim to analyze the impact of different weather events on 
public
health and economy based on the storm database collected from the U.S. National 
Oceanic and Atmospheric Administration's (NOAA) from 1950 - 2011. We will use the
estimates of fatalities, injuries, property and crop damage to decide which types
of event are most harmful to the population health and economy. From these data, 
we found that excessive heat and tornado are most harmful with respect to 
population,health, while flood, drought, and hurricane/typhoon have the greatest 
economic consequences.

## Data Processing

The raw data were available from a URL provided by the Coursera project 
instructions, and were downloaded using the download.file() function in R:

```{r, echo = TRUE}

library(knitr)
library(markdown)
library(rmarkdown)
library(plyr)
library(stats)

download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.c
sv.bz2","storms.csv.bz2")

```

The data were then read into R using the read.csv() function:

```{r, echo=TRUE}
storms <- read.csv("storms.csv.bz2",stringsAsFactors=F)
dim(storms)
```

```{r, echo=TRUE}

names(storms)

```

```{r,echo=TRUE}
str(storms)
```

There are 902.297 observations with 37 variables in the file. Only a subset is 
requrired for the analysis.  
1. Relevant for the analysis are the date (BGN_DATE), event type (EVTYPE), 
counter for the health impact (FATALITIES and INJURIES), monetary impact on crop 
and property (PROPDMG and CROPDMG) as well as their corresponding exponents 
(PROPDMGEXP and CROPDMGEXP).  
2. According to the NOAA (https://www.ncdc.noaa.gov/stormevents/details.jsp) the 
full set of wheather events (48 event types) is available since 1996. Between 
1950 and 1995 only a subset (Tornado, Thunderstorm Wind and Hail) of these events
is available in the storm database. In order to have o comparable basis for the 
analysis the dataset is limited to the observations between 1996 and 2011.  
3. The dataset contains a lot of observations without any information about 
health and/or economic damages. These observations are excluded from the 
analysis.

## Wrangling the Data

Defining variables that will be used:

EVTYPE: Event Type (Tornados, Flood, ..)

FATALITIES: Number of Fatalities

INJURIES: Number of Injuries

PROGDMG: Property Damage

PROPDMGEXP: Units for Property Damage (magnitudes - K,B,M)

CROPDMG: Crop Damage

CROPDMGEXP: Units for Crop Damage (magnitudes - K,BM,B)

```{r, echo = TRUE}
varsNeeded <- c("EVTYPE", "FATALITIES", "INJURIES", "PROPDMG", "PROPDMGEXP", 
"CROPDMG", "CROPDMGEXP")
storms <- storms[varsNeeded]
dim(storms)
```

```{r, echo=TRUE}
names(storms)
```

## Total for Property Damage

Refactor of variable PROPDNGEXP
```{r, echo=TRUE}
unique(storms$PROPDMGEXP)

```

```{r,echo=TRUE}
storms$PROPDMGEXP <- mapvalues(storms$PROPDMGEXP, from = c("K", "M","", "B", "m",
"+", "0", "5", "6", "?", "4", "2", "3", "h", "7", "H", "-", "1", "8"), to = 
c(10^3, 10^6, 1, 10^9, 10^6, 0,1,10^5, 10^6, 0, 10^4, 10^2, 10^3, 10^2, 10^7, 
10^2, 0, 10, 10^8))
storms$PROPDMGEXP <- as.numeric(as.character(storms$PROPDMGEXP))
storms$PROPDMGTOTAL <- (storms$PROPDMG * storms$PROPDMGEXP)/1000000000
```

Refactor of variable CROPDMGEXP variable
```{r,echo=TRUE}
unique(storms$CROPDMGEXP)
```

```{r, echo=TRUE}
storms$CROPDMGEXP <- mapvalues(storms$CROPDMGEXP, from = c("","M", "K", "m", "B",
"?", "0", "k","2"), to = c(1,10^6, 10^3, 10^6, 10^9, 0, 1, 10^3, 10^2))
storms$CROPDMGEXP <- as.numeric(as.character(storms$CROPDMGEXP))
storms$CROPDMGTOTAL <- (storms$CROPDMG * storms$CROPDMGEXP)/1000000000
```

## Results

# Events that are Most Harmful to Population Health

To determine which type of events are most harmful to the population health we
must look at the variables fatalities and Injuries.

# 1. Fatalities

```{r, echo=TRUE}

sumFatalities <- aggregate(FATALITIES ~ EVTYPE, data = storms,  FUN="sum")
dim(sumFatalities)
```


Ordering Number of Fatalities and defining the top 10 Weather events

```{r, echo=TRUE}
fatalities10events <- sumFatalities[order(-sumFatalities$FATALITIES), ][1:10, ]
dim(fatalities10events)
```

```{r, echo= TRUE}
fatalities10events
```

BarPlot of the 10 Fatalities Events most harmful to population Health.

```{r, echo= TRUE}

par(mfrow = c(1,1), mar = c(12, 4, 3, 2), mgp = c(3, 1, 0), cex = 0.8)
barplot(fatalities10events$FATALITIES, names.arg = fatalities10events$EVTYPE, las
= 3, main = "10 Fatalities Highest Events", ylab = "Number of Fatalities")
```


```{r, echo = TRUE}
dev.copy(png, "fatalities-events.png", width = 480, height = 480)
```

```{r, echo= TRUE}
dev.off()
```

# 2. Injuries

Number of Injuries per type of Event (EVTYPE)

```{r, echo= TRUE}
sumInjuries <- aggregate(INJURIES ~ EVTYPE, data = storms,  FUN="sum")
dim(sumInjuries)
```

Ordering Number of INJURIES and defining the top 10 Weather events in this 
category

```{r, echo= TRUE}
injuries10events <- sumInjuries[order(-sumInjuries$INJURIES), ][1:10, ]
dim(injuries10events)
```

```{r, echo= TRUE}
injuries10events
```

BarPlot of the 10 INJURIES Events most harmful to population Health.

```{r. echo = TRUE}
par(mfrow = c(1,1), mar = c(12, 6, 3, 2), mgp = c(4, 1, 0), cex = 0.8)
barplot(injuries10events$INJURIES, names.arg = injuries10events$EVTYPE, las = 3, 
main = "10 Injuries Highest Events", ylab = "Number of Injuries")
```


```{r, echo= TRUE}
dev.copy(png, "injuries-events.png", width = 480, height = 480)
```

```{r,echo= TRUE}
dev.off()
```
# Events that have the Greatest Economic Consequences

To determine which type of events have the greatest econimic consequences the 
variables, PROPDMG (Property Damage) and CROPDMG (Crop Damage) have to be taken
into consideration.

# 1.Property Damage

Calculation of property Damage

```{r, echo= TRUE}
sumPropertyDamage <- aggregate(PROPDMGTOTAL ~ EVTYPE, data = storms,  FUN="sum")
dim(sumPropertyDamage)
```

Top 10 highest Property damage Events

```{r, echo= TRUE}
propdmg10Total <- sumPropertyDamage[order(-sumPropertyDamage$PROPDMGTOTAL), 
][1:10, ]
propdmg10Total
```

BarPlot of the top 10 events most harmful to population economic.

```{r, echo= TRUE}
par(mfrow = c(1,1), mar = c(12, 6, 3, 2), mgp = c(3, 1, 0), cex = 0.8)
barplot(propdmg10Total$PROPDMGTOTAL, names.arg = propdmg10Total$EVTYPE, las = 3, 
main = "10 Property Damages Highest Events", ylab = "Damage Property Values (in 
Billions)")
```

```{r, echo= TRUE}
dev.copy(png, "propdmg-total.png", width = 480, height = 480)
```

```{r, echo= TRUE}
dev.off()
```

# 2. Crop Damage

```{r, echo= TRUE}
sumCropDamage <- aggregate(CROPDMGTOTAL ~ EVTYPE, data = storms,  FUN="sum")
dim(sumCropDamage)
```

Top 10 highest crop damage events

```{r, echo= TRUE}
cropdmg10Total <- sumCropDamage[order(-sumCropDamage$CROPDMGTOTAL), ][1:10, ]
cropdmg10Total
```

BarPlot of the 10 Crop Damage Events most harmful to population economic.

```{r, echo= TRUE}
par(mfrow = c(1,1), mar = c(10, 6, 3, 2), mgp = c(3, 1, 0), cex = 0.6)
barplot(cropdmg10Total$CROPDMGTOTAL, names.arg = cropdmg10Total$EVTYPE, las = 2, 
main = "10 Crop Damages Highest Events", ylab = "Damage Crop Values (in Billions)
")
```

```{r, echo= TRUE}
dev.copy(png, "cropdmg-total.png", width = 480, height = 480)
```

```{r, echo= TRUE}
dev.off()
```

## Final Notes
It should be noted that the event rates computed here do not take account of two 
important factors: time and area. A HEATWAVE event usually takes place over 
several days and in a large area, as do cold weather events, tropical storms, 
hurricances, and others. Some events may be unlisted in the rankings because they
take place over a shorter time period and/or act over a smaller area, putting 
fewer people and less property at risk. A more thorough analysis should take 
these factors into account, potentially by computing the rankings within a higher
level of weather event category, over geographic size categories, and by 
duration. As always, asking the right question and understanding the data are 
paramount to reaching good conclusions.
